---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Shatavia Bellmon"
date: "3/2/26"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r}
staff <- read_csv("data/instructional-staff.csv")
```
```{r}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
```

```{r}
staff_long %>% head()
```

```{r}
staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()
```

### Exercise 1

```{r}
ggplot(staff_long, aes(x = year, y = value, group = faculty_type, color = faculty_type)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Instructional Staff Employment Trends",
    x = "Year",
    y = "Percentage of Hires",
    color = "Faculty Type"
  ) +
  theme_minimal()
```

### Exercise 2

```{r}
staff_long <- staff_long %>%
  mutate(highlight = if_else(faculty_type == "Part-Time Faculty", "Part-Time", "Other"))
```

```{r}
ggplot(staff_long, aes(x = year, y = value, group = faculty_type, color = highlight)) +
  geom_line(aes(size = highlight)) +
  scale_color_manual(values = c("Part-Time" = "red", "Other" = "blue")) +
  scale_size_manual(values = c("Part-Time" = 1.5, "Other" = 0.5)) +
  labs(
    title = "Proportion of Part-Time Faculty Over Time",
    x = "Year",
    y = "Percentage of Hires",
    color = "Legend"
  ) +
  theme_minimal()
```

# I would have reshaped the data from wide to long format. I would also choose a line graph because it shows continuity and trends over time better than the other plot used. I would also add different colors to the different lines to ensure that differentiation across trends was easy to see.

# In the updated plot, I created a new variable to "highlight" the part-time faculty category in a bold color red, while putting the other faculty types in blue. I also increased the line thickness for the part-time category. This makes sure all lines are seen clearly, allowing the viewer to immediately see the different trends. 


### Exercise 3

```{r}
fisheries <- read_csv("data/fisheries.csv")
```
# I would replace the pie charts with faceted bar charts to make the data easier to read.
# I would flip the chart as well just because there are lot of countries represented. Each piece of data should be seen clearly.
# Another suggestion could be grouping the countries by continents to look at regional trends.

View(fisheries)


```{r}
# Basic Horizontal Plot
ggplot(fisheries, aes(x = country, y = total)) +
  geom_col() +
  coord_flip() + 
  labs(
    title = "Fishery Production",
    x = "Country",
    y = "Percentage of Total Tonnage"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 2))
```


```{r}
library(tidyverse)
library(countrycode)


fisheries_improved <- fisheries %>%
  mutate(
    # Created continent groups
    continent = countrycode(country, origin = "country.name", destination = "continent"),
    continent = if_else(is.na(continent), "Other/Unknown", continent),
    country = reorder(country, total) 
  )

# visualization 
ggplot(fisheries_improved, aes(x = country, y = total, fill = continent)) +
  geom_col() +
  coord_flip() + 
  facet_wrap(~ continent, scales = "free_y") + 
  labs(
    title = "Fishery Production by Country and Region",
    subtitle = "Faceted by continent to show regional trends",
    x = "Country",
    y = "Percentage of Total Capture",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 2))
```

# I replaced the original 3D pie charts with faceted bar charts because bar charts allow for a more accurate linear comparison of values, whereas 3D pie charts could distort the data's proportions. I used the coord_flip() function to create a horizontal layout, ensuring that the names of the many countries represented are legible and do not overlap. I grouped countries by continent to highlight regional production trends that were previously hidden in the original visualization.

